---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# auk: eBird Data Processing with AWK

[![Licence](https://img.shields.io/github/license/mashape/apistatus.svg)](http://choosealicense.com/licenses/mit/)
[![Travis-CI Build Status](https://img.shields.io/travis/mstrimas/auk/master.svg?label=Mac%20OSX%20%26%20Linux)](https://travis-ci.org/mstrimas/auk)
[![AppVeyor Build Status](https://img.shields.io/appveyor/ci/mstrimas/auk/master.svg?label=Windows)](https://ci.appveyor.com/project/mstrimas/auk)
[![Coverage Status](https://img.shields.io/codecov/c/github/mstrimas/auk/master.svg)](https://codecov.io/github/mstrimas/auk?branch=master)
[![CRAN\_Status\_Badge](http://www.r-pkg.org/badges/version/auk)](https://cran.r-project.org/package=auk)

**This package is the early stages of development and should be used with caution. If you encounter any bugs, please open an issue on GitHub**

## Overview

[eBird](http://www.ebird.org) is an online tool for recording bird observations. Since its inception, nearly 500 million sightings records have been collected, making it one of the largest citizen science projects in history and an extremely valuable resource for bird research and conservation. The full eBird database is packaged as a text file and available for download as the [eBird Basic Dataset (EBD)](http://ebird.org/ebird/data/download). Due to the large size of this dataset, it must be filtered to a smaller subset of desired observations before reading into R. This filtering is most efficiently done using AWK, a Unix utility and programming language for processing column formatted text data. This package acts as a front end for AWK, allowing users to filter eBird data before import into R.

## Installation

This package can be installed directly from GitHub with: 

```{r gh-install, eval=FALSE}
# install.packages("devtools")
devtools::install_github("mstrimas/auk")
```

`auk` requires the Unix utility AWK, which is available on most Linux and Mac OS X machines. Windows users will first need to install [Cygwin](https://www.cygwin.com) before using this package.

## Vignette

Full details on using `auk` to produce both presence-only and presence-absence data are outlined in the vignette, which can be accessed with `vignette("auk")`.

## Usage

### Cleaning

Some rows in the eBird Basic Dataset (EBD) may have an incorrect number of columns and the dataset has an extra blank column at the end. The function `auk_clean()` drops these erroneous records and removes the blank column.

```{r auk-clean}
library(auk)
# sample data
f <- system.file("extdata/ebd-sample_messy.txt", package = "auk")
tmp <- tempfile()
# remove problem records
auk_clean(f, tmp)
# number of lines in input
length(readLines(f))
# number of lines in output
length(readLines(tmp))
unlink(tmp)
```

### Filtering

`auk` uses a [pipeline-based workflow](http://r4ds.had.co.nz/pipes.html) for defining filters, which can be compiled into an AWK script. Users should start by defining a reference to the EBD file with `auk_ebd()`. Then any of the following filters can be applied:

- `auk_species()`: filter by species using common or scientific names.
- `auk_country()`: filter by country using the standard English name or [ISO 2-letter country codes](https://en.wikipedia.org/wiki/ISO_3166-1_alpha-2).
- `auk_extent()`: filter by spatial extent, i.e. a range of latitudes and longitudes.
- `auk_date()`: filter to checklists from a range of dates.
- `auk_last_edited()`: filter to checklists from a range of last edited dates, useful for extracting just new or recently edited data.
- `auk_time()`: filter to checklists started during a range of times.
- `auk_duration()`: filter to checklists that lasted a given length of time.
- `auk_complete()`: only retain checklists in which the observer has specified that they recorded all species seen or heard. These records are the most useful for modelling because they provide both presence and absence data.

```{r auk-filter}
# sample data
f <- system.file("extdata/ebd-sample.txt", package = "auk")
# define an EBD reference and a set of filters
ebd <- auk_ebd(f) %>% 
  # species: common and scientific names can be mixed
  auk_species(species = c("Gray Jay", "Cyanocitta cristata")) %>%
  # country: codes and names can be mixed; case insensitive
  auk_country(country = c("US", "Canada", "mexico")) %>%
  # extent: formatted as `c(lng_min, lat_min, lng_max, lat_max)`
  auk_extent(extent = c(-100, 37, -80, 52)) %>%
  # date: use standard ISO date format `"YYYY-MM-DD"`
  auk_date(date = c("2012-01-01", "2012-12-31")) %>%
  # time: 24h format
  auk_time(time = c("06:00", "09:00")) %>%
  # duration: length in minutes of checklists
  auk_duration(duration = c(0, 60)) %>%
  # complete: all species seen or heard are recorded
  auk_complete()
ebd
```

In all cases, checks are performed to ensure filters are valid. For example, species are checked against the official [eBird taxonomy](http://help.ebird.org/customer/portal/articles/1006825-the-ebird-taxonomy) and countries are checked using the [`countrycode`](https://github.com/vincentarelbundock/countrycode) package.

Each of these functions only defines the filter. `auk_filter()` should be used to compile all the filters into an AWK script and execute it to produce an output file. So, bringing all this together, one could, for example, extract all Gray Jay and Blue Jay records from Canada with:

```{r auk-complete, eval = FALSE}
tmp <- tempfile()
ebd <- system.file("extdata/ebd-sample.txt", package = "auk") %>% 
  auk_ebd() %>% 
  auk_species(species = c("Gray Jay", "Cyanocitta cristata")) %>% 
  auk_country(country = "Canada") %>% 
  auk_filter(file = tmp)
unlink(tmp)
```

### Reading

EBD files can be read with `read_ebd()`:

```{r read}
system.file("extdata/ebd-sample.txt", package = "auk") %>% 
  read_ebd() %>% 
  str()
```

By default, `read_ebd()` returns a tibble for use with [Tidyverse](http://tidyverse.org) packages. Tibbles will behave just like plain data frames in most instances, but users can choose to return a plain `data.frame` or `data.table` by using the `setclass` argument.

```{r read-tbl}
ebd_df <- system.file("extdata/ebd-sample.txt", package = "auk") %>% 
  read_ebd(setclass = "data.frame")
```


## Acknowledgements

This package is based on the AWK scripts provided in a presentation given by Wesley Hochachka, Daniel Fink, Tom Auer, and Frank La Sorte at the 2016 NAOC eBird Data Workshop on August 15, 2016.

## References

```
eBird Basic Dataset. Version: ebd_relFeb-2017. Cornell Lab of Ornithology, Ithaca, New York. May 2013.
```
